Reasoning about Connectors Using Coq and Z3

### 2017-06-24 FACS_Extension/paper_revised0508.tex
paper draft finished

%\documentclass[3p,times]{elsarticle}
%\usepackage{ecrc}
\documentclass[preprint,3p]{elsarticle}
\usepackage{graphicx}
% used for flow charts
\usepackage{listings}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{reotex}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

@@

\lstset{
    basicstyle=\ttfamily,
    numbers=left,
    numberstyle=\small,
    columns=flexible,
    numbersep=10pt,
    tabsize=2,
    extendedchars=true,         %
    breaklines=true,
    keywordstyle=\bfseries,
    stringstyle=\color{white}\ttfamily, % Farbe der String
    xleftmargin=20pt,
    framexleftmargin=17pt,
    framexrightmargin=5pt,
    framexbottommargin=4pt,
    showstringspaces=false,
    frame=single
}

\lstdefinelanguage{coq}{
    keywords = {
        Parameter, Parameters, Variable, Variables,
        Fixpoint,
        Stream, Prop,
        forall, exists,
        Axiom, Lemma, Goal, Theorem,
        Proof, Qed
    }
}

@@

\title{Compositional Reasoning about Connectors Using Coq and Z3}

@@

%\maketitle
\begin{abstract}
Reo is a channel-based exogenous coordination language in which complex coordinators, called connectors, are compositionally built out of simpler ones. In this paper, we present an approach to model and reason about connectors using Coq and Z3. %Coq is a proof assistant based on higher-order logic and $\lambda$-calculus.
The model reflects the original structure of connectors simply and clearly. In our framework, basic connectors (channels) are interpreted as axioms and composition operations are specified as inference rules. Furthermore, connectors are interpreted as logical predicates which describe the relations
between inputs and outputs. With such definitions provided, connector properties, as well as equivalence and refinement relations between different connectors, can be naturally formalized as \emph{goals} in Coq and easily proved using pre-defined \emph{tactics}. Particularly, to fulfill the proof of refinement relation, Z3 is used as an assistant tool for solving logical formulas.
%\paragraph*{Keywords:}

@@

In the current paper, we aim to provide a more complete account of the approach in \cite{ZHL+17} to formally model and reason about connectors. The main drawback of failing to check the complementary refinement relation in Coq is made up by using the constraint solver Z3. While Coq provides a platform to prove properties with a series of tactics and strategies to be considered, Z3 can be used as a component in the context %of other tools
that requires solving logical formulas. The logical formulas involved in the model of Reo should reflect the constraints of connectors and be consistent with their behavior.

The approach is not a brand new idea, as we have already provided a solution for modeling Reo in Coq in \cite{LS15}, where connectors are represented in a constructive way, and verification is essentially based on simulations. We do believe that the approach in this paper is reasonably different from its predecessor \cite{LS15} where Coq seldom shows its real power. To be more specific, our new work has its certain
advantages comparing with \cite{LS15} in the following aspects:
\begin{itemize}
\item {\bf Modeling Method:} We use axioms to describe basic channels and their composition operations, which is more natural on a proof-assistant platform than the simulation-based approach in \cite{LS15}.
\item {\bf Expression Power:} Any valid Coq expression can be used to depict properties, which is obviously more powerful than just using LTL formulas in \cite{LS15}. Furthermore, support for continuous time behavior is also possible in our approach in this paper.
\item {\bf Refinement and Equivalence Checking:} In our framework, equivalence and refinement relations can be proved among different connectors, while the approach in \cite{LS15} is not capable of either equivalence or refinement checking. Furthermore, the constraint solver Z3 is adopted to prove non-refinement relations between connectors, while in \cite{ZHL+17} we can only derive proofs for refinement relations.
\end{itemize}

The paper is organized as follows: After this general introduction, we briefly summarize Reo, Coq and Z3
in Section \ref{sec:pre}. Section \ref{sec:basicdef} shows the notion of timed data streams and some
pre-defined auxiliary functions and predicates. Section \ref{sec:channelandoperator} presents the
formal modeling of basic channels and operators, as well as complex connectors. Section
\ref{sec:verification} shows how to reason about connector properties and equivalence (or
refinement) relations in our framework. Section \ref{sec:refinement} concentrates on the proof of refinement relation between two connectors.
In Section \ref{sec:conclusion}, we conclude with some further research directions. Full source codes can be found at \cite{reo2coqfile} for further reference.

@@

\subsection{The Coordination Model Reo}
Reo is a channel-based exogenous coordination language wherein complex coordinators, called connectors,
are compositionally built out of simpler ones \cite{Arb04}.
Further details about Reo and its semantics can be found in \cite{Arb04, AR03, BSAR06}.
The simplest connectors are channels with well-defined behavior such as synchronous channels, FIFO channels, etc.
Each channel in Reo has exactly two directed ends, with their own identities.
There are two types of channel ends: \emph{source} ends and \emph{sink} ends. A source channel end accepts
data into the channel. A sink channel end dispenses data out of the channel.
\begin{figure}
  \centering
  \begin{tikzpicture}
  \draw (0.7, 0.6) node[draw=none] {Sync};
  \sync{(0,1)}{(1.5,1)}{}
  \draw (2.7, 0.6) node[draw=none] {Syncdrain};
  \syncdrain{(2,1)}{(3.5,1)}{node[below]{}}
  \draw (4.7, 0.6) node[draw=none] {FIFO1};
  \fifoe{(4,1)}{(5.5,1)}{node[below]{}}
  \draw (6.7, 0.6) node[draw=none] {LossySync};
  \lossysync{(6,1)}{(7.5,1)}{node[below left]{}}
  \end{tikzpicture}\vspace{-6mm}
  \caption{Four types of basic channels.}\label{fig:basicchannel}
\end{figure}

The graphical notations of some basic channels are presented in Fig.~\ref{fig:basicchannel}, and their behavior can be interpreted as follows:
\begin{itemize}
\item{\textbf{Sync:} a synchronous channel with one source end and one sink end. The pair of I/O operations on its two ends can succeed only simultaneously.}
\item{\textbf{SyncDrain:} a synchronous channel which has two source ends.
 The pair of input operations on its two ends can succeed only simultaneously.
 All data items written to this channel are lost.}
\item{\textbf{FIFO$n$:} an asynchronous channel with one source end and one sink end, and a bounded buffer with capacity $n$.
It can accept data items from its source end. The accepted data items are kept in the internal buffer, and dispensed to
the sink end in FIFO order.
Especially, the FIFO$1$ channel is an instace of FIFO$n$ where the buffer capacity is 1.}
%\item{\textbf{AsyncDrain:}  an asynchronous channel} which has two source ends. The channel guarantees that the operations on its two ends never succeed simultaneously. All data items written to this channel are lost.
\item{\textbf{LossySync:} a synchronous channel with one source end
    and one sink end. The source end always accepts all data items. If
    there is no matching output operation on the sink end of the
    channel at the time that a data item is accepted, then the data
    item is lost; otherwise, the channel transfers the data item
    exactly the same as a Sync channel, and the output operation at the sink end succeeds.}
\end{itemize}

@@

Complex connectors are constructed by composing simpler ones via the join and hiding operations. Channels are joined together in nodes. The set of channel ends coincident on a node is disjointly partitioned into the sets of source and sink channel ends that coincide on the node, respectively. Nodes are categorized into source, sink and mixed nodes, depending on whether all channel ends that coincide on a node are source ends, sink ends or a combination of the two. The hiding operation is used to hide the internal topology of a connector. The hidden nodes can no longer be accessed or observed from outside. There are three types of operations for channel composition: \emph{flow-through}, \emph{merge} and \emph{replicate}. Fig.~\ref{fig:channelcomposition} provides the graphical representation of these operations.

\subsection{The Proof Assistant tools}

Coq\cite{huet1997coq} is a widely-used proof assistant tool, where denotational formalizations (e.g.
theorem and hypothesis) and operational formalizations (e.g. functions and algorithms) are naturally
integrated. Moreover, it allows the interactive construction of formal proofs.
The formal language used in Coq, \emph{Gallina}, provides a convenient way to define
both programming statements and mathematical propositions, for example:
\begin{lstlisting}[language=coq]
  (* a variable definition *)
  Variables a b: nat.
  (* a simple non-recursive function *)
  Definition inc(a:nat) := a + 1.
  (* axioms don't have to be proved *)
  Axiom inc_ax: forall c:nat, inc(c) > c.
  (* theorems rely on proving *)
  Theorem inc_eq: forall c:nat, inc(c) = c + 1.
  Proof.
    (* interactive proving based on tactics *)
    auto.
  Qed.
\end{lstlisting}

@@

Furthermore, Coq is equipped with a set of well-written standard libraries. For example, as used in
this paper, \emph{List} describes the widely-used finite list structure, \emph{Stream} provides a co-inductive
definition of infinite lists, and \emph{Reals} defines various operations and theorems on real
numbers. Usually, quite a few lemmas and theorems are pre-defined in such libraries, making it
substantially easier to prove our goals.

@@

Z3 \cite{MouraB08} is an efficient SMT (Satisfiability Modulo Theories) Solver and a high performance theorem prover freely available from Microsoft. It can be used in various software verification and analysis applications. Acting as a SMT solver, Z3 expands to deciding the satisfiability (or dually the validity) of first order formulas with respect to combinations of theories such as: arithmetic, bit-vectors, arrays, and uninterpreted functions. Given the data and time constraints of connectors, it allows us to decide the satisfiability against properties or refinement relations. Z3 provides bindings for several programming languages. In this paper, we use the Z3 API in Python to construct the models and carry out experiments. The following example provides an intuitive understanding of the Z3 solver.
\begin{lstlisting}[frame=single]
x = Int('x')
y = Int('y')

s = Solver()
s.add(x > 10, Or(x + y > 3, x - y < 2))
#Solving constraints in the solver s
print s.check()

#Create a new scope
s.push()
s.add(y < 11)
#Solving updated set of constraints
print s.check()

#Restoring state
s.pop()
#Solving restored set of constraints
print s.check()
\end{lstlisting}

This example contains two constraints at first. From these two constraints, we can see that Z3 supports Boolean operators, such as \emph{And}, \emph{Or}, \emph{Not} and \emph{Implies (implication)}, apart from the operators like $<$, $>$ for comparison.
The command \emph{Solver()} creates a general purpose solver. Constraints can be added using the method \emph{add}. Once they are added to the solver, we can say that the constraints have been asserted in the solver. The method \emph{check()} solves the asserted constraints. Finally, the result is satisfiable (unsatisfiable) if a solution is (not) found, respectively. A solver may fail to solve a system of constraints and \emph{unknown} is returned. Each solver maintains a stack of assertions. The command \emph{push} creates a new scope by saving the current stack size. The command \emph{pop} removes any assertion performed between it and the matching \emph{push}.
\section{Basic Definitions}\label{sec:basicdef}

In this section, we briefly introduce the notion of timed data streams and some pre-defined auxiliary functions and predicates in Coq, which are used in the following sections for modeling connectors.

@@

%\subsection{Auxiliary functions and predicates}
Some auxiliary functions and predicates are defined to facilitate the representation of axioms for basic channels in Reo. This part can be extended for further use in different problems.
The terms ``PrL" and ``PrR" take a pair of values (a, b) that has Cartesian product type A $\times$ B as the argument and return the first or second value of the pair, respectively.
%\begin{verbatim}
%    Definition PrL {A B: Type}(pair: A * B) := (...).
%    Definition PrR {A B: Type}(pair: A * B) := (...).
%\end{verbatim}
The following functions provide some judgment of time, which can make
the description of axioms and theorems for connectors more concise and
clear: ``Teq" means that time of two streams are equal and ``Tneq" has the
opposite meaning. ``Tle"  (``Tgt") represents that time of the first stream is
strictly less (greater) than the second stream. The judgement about
equality of data is analogous to the judgement of time. The complete
definition of these functions can be found at \cite{reo2coq2Z3}.

\section{Formal Modeling of Basic Channels and Operators}\label{sec:channelandoperator}
In this section, we show how primitive connectors, i.e., channels, and operators for connector composition are specified in Coq and used for modeling of complex connectors. Then we can apply the tactics provided in Coq to reason about connector properties. Basic channels, which can be regarded as axioms of the whole framework, are specified as logical predicates illustrating the relation between the timed data streams of input and output. When we need to construct a more complex connector, appropriate composition operators are applied depending on the topological structure of the connector.

@@

The channel types FIFO and FIFO$n$ where $n$ is an integer greater than $0$ represent the
typical unbounded and bounded asynchronous FIFO channels. A write to a FIFO channel always succeeds, and a write to a FIFO$n$ channel succeeds only if the
number of data items in its buffer is less than its bounded capacity $n$. A read or take from a FIFO or FIFO$n$ channel suspends until the first data item
in the channel buffer can be obtained and then the operation succeeds.
%We use the $\overrightarrow{\emph{a}}$ to denote the tail of  a sequence \emph{a}.
For simplicity, we take the FIFO1 channel as an example. This channel type requires that the time when it consumes a data item through its source end is earlier than the time when the data item is delivered through its sink end. Besides, as the buffer has the capacity 1, time of the next data item that flows in should be later than the time when the data in the buffer is delivered. We use intersection of predicates in its definition as follows:
\begin{lstlisting}[language=coq]
    Definition FIFO1(Input Output:Stream TD) : Prop :=
    Tle Input Output /\ Tle Output (tl Input) /\ Deq Input Output.
\end{lstlisting}

For a FIFO1 channel whose buffer already contains a data element \emph{e}, the communication can be initiated only if the data element \emph{e} can be taken via the sink end. In this case, the data stream that flows out of the channel should get an extra element \emph{e} settled at the beginning of the stream. And time of the stream that flows into the channel should be earlier than time of the tail of the stream that flows out. But as the buffer contains the data element \emph{e}, new data can be written into the channel only after the element \emph{e} has been taken. Therefore, time of the stream that flows out is earlier than time of the stream that flows in. The channel can be represented as the \begin{lstlisting}[language=coq]
    Definition FIFO1e(Input Output:Stream TD)(e:Data) : Prop :=
    Tgt Input Output /\ Tle Input (tl Output)
    /\ PrR (hd Output) = e  /\ Deq Input (tl Output).
\end{lstlisting}

In the following we choose Axiom to define LossySync and AsyncDrain
because it is easier to use the coinductive expression to specify
their behavior.

A LossySync channel behaves the same as a Sync channel, except that a write operation on its source end always succeeds immediately. If a compatible read or take operation is already pending on its sink end, the written data item is transferred to the pending operation and both succeed. Otherwise, the write operation succeeds and the data item is lost. The LossySync channel can be defined as follows:
\begin{lstlisting}[language=coq]
   Parameter LossySync: Stream TD -> Stream TD -> Prop.
   Axiom LossySync_coind:
     forall Input Output: Stream TD,
     LossySync Input Output ->
       (
        (hd Output = hd Input  /\ LossySync (tl Input)(tl Output)) \/
        LossySync(tl Input) Output
       ).
\end{lstlisting}

AsyncDrain is analogous to SyncDrain except that it guarantees that the pairs of write operations on the two channel ends never succeed simultaneously. Similarly it only has requirements on the time of the two streams on its opposite ends, but it requires that the times of the two streams are always different. The AsyncDrain channel can be defined as follows:
\begin{lstlisting}[language=coq]
    Parameter AsyncDrain: Stream TD -> Stream TD -> Prop.
    Axiom AsyncDrain_coind:
      forall Input1 Input2: Stream TD,
      AsyncDrain Input1 Input2 ->
        (~ PrL(hd Input1)  =  PrL (hd Input2) ) /\
        (
          ( (PrL(hd Input1) < PrL (hd Input2)) /\ AsyncDrain (tl Input1) Input2) /\
          ( (PrL(hd Input1) > PrL (hd Input2))  /\ AsyncDrain Input1 (tl Input2) )
        ).
\end{lstlisting}

@@

Based on the definition of basic channels and operators, more complex connectors can be constructed structurally.
%As the whole structure of each connector has been determined, the rest work is to use the operators to connect appropriate  channels according to the structure of each connector.
To show how a composite connector is constructed, we consider a simple example as shown in Fig. \ref{fig:compsyncfifo}, where a FIFO1 channel is attached
to the sink end of a Sync channel. Assume \emph{AB} is of type Sync and \emph{BC} is of type FIFO1, then we can construct the required connector by
illustrating \emph{Sync(A,B)} and \emph{FIFO1(B,C)}. The configuration and the functionality of the required connector can be specified using this concise method. Note that the composition operations can be easily generalized to the case of multiple nodes, where the modeling of connectors is similar. More
examples can be found in Section \ref{sec:verification}.

\section{Reasoning about Connectors}\label{sec:verification}
After modeling a connector in Coq, we can analyse and prove important properties of the connector. In this section, we give some examples to elucidate how to
reason about connector properties and prove refinement/equivalence relations between different connectors, with the help of Coq.

\subsection{Derivation of Connector Properties}
The proof process of a property is as follows: the user states the proposition that needs to be proved, called a \emph{goal},
then he/she applies commands called \emph{tactics} to decompose this goal into simpler subgoals or solve it directly. This decomposition
process ends when all subgoals are completely solved. In the following, we use some examples to illustrate our approach instead of
giving all the complex technical details.
\begin{example}
We first consider the connector given in Fig. \ref{fig:compsyncfifo}, which consists of two channels \emph{AB} and \emph{BC} with types Sync and FIFO1, respectively.
We use $a$ and $b$ to denote the time streams when the corresponding data streams flow into and out of the Sync channel \emph{AB}, and
$c$ to denote the time stream for the data stream that flows out of the FIFO1 channel \emph{BC}. Here we can see that a flow-through
operation has acted on the mixed node \emph{B}. The time when the stream flows into the FIFO1 channel \emph{BC} is equal to the time when the
stream flows out of the Sync channel \emph{AB}. The following theorem states the property $a < c$ for this connector. The connector is based on the
axioms Sync and FIFO1, which can be used as hypotheses for the proof of the following theorem.
\end{example}

\begin{theorem}\label{the:tleac}
$\forall A,B,C.\:Sync(A,B)\land FIFO1(B,C) \rightarrow Tle(A,C)$.
\end{theorem}

>>>>

%\documentclass[3p,times]{elsarticle}
%\usepackage{ecrc}
\documentclass[preprint,3p]{elsarticle}
\usepackage{graphicx}
% used for flow charts
\usepackage{listings}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{reotex}
\usepackage{array}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

@@

\lstset{
    basicstyle=\ttfamily\small,
    numbers=left,
    numberstyle=\small,
    columns=flexible,
    numbersep=10pt,
    tabsize=2,
    extendedchars=true,         %
    breaklines=true,
    keywordstyle=\bfseries,
    stringstyle=\color{white}\ttfamily, % Farbe der String
    xleftmargin=20pt,
    framexleftmargin=17pt,
    framexrightmargin=5pt,
    framexbottommargin=4pt,
    showstringspaces=false
    % frame=single
}

\lstdefinelanguage{coq}{
    keywords = {
        Parameter, Parameters, Variable, Variables,
        Fixpoint,
        Definition,
        Stream, Prop,
        forall, exists,
        Axiom, Lemma, Goal, Theorem,
        Proof, Qed,
        Hypothesis, Section,
    }
}

@@

\title{Reasoning about Connectors Using Coq and Z3}

@@

%\maketitle
\begin{abstract}
Reo is a channel-based exogenous coordination language in which complex coordinators, called connectors, are compositionally built out of simpler ones. In this paper, we present an approach to model and reason about connectors using Coq and Z3.
The model reflects the original structure of connectors simply and clearly. In our framework, both basic connectors (channels) and composition operations are interpreted as axioms. Furthermore, connectors are interpreted as logical predicates which describe the relations
between inputs and outputs. With such definitions provided, connector properties, as well as equivalence and refinement relations between different connectors, can be naturally formalized as \emph{goals} in Coq and proved using pre-defined \emph{tactics}, if satisfied by connectors. Furthermore, when fail to prove the properties, we use Z3, the SMT solver, to search for possible bounded counter examples automatically.
% Particularly, to fulfill the proof of refinement relation, Z3 is used as an assistant tool for solving logical formulas.
%\paragraph*{Keywords:}

@@

In the current paper, we aim to provide a more complete account of the approach in \cite{ZHL+17} to formally model and reason about connectors. The main drawback of failing to check the complementary refinement relation in Coq is partially made up by using the constraint solver Z3. While Coq provides a platform to prove properties with a series of tactics and strategies to be considered, Z3 can be used as a component in the context %of other tools
that requires solving logical formulas. The logical formulas involved in the model of Reo should reflect the constraints of connectors and be consistent with their behavior.

The approach is not a brand new idea, as we have already provided a solution for modeling Reo in Coq in \cite{LS15}, where connectors are represented in a constructive way, and verification is essentially based on simulations. We do believe that the approach in this paper is reasonably different from its predecessor \cite{LS15} where Coq seldom shows its real power. To be more specific, our new work has its certain
advantages comparing with \cite{LS15} in the following aspects:
\begin{itemize}
\item {\bf Modeling Method:} We use axioms to describe basic channels and their composition operations, which is more natural on a proof-assistant platform than the simulation-based approach in \cite{LS15}.
\item {\bf Expression Power:} Any valid Coq expression can be used to depict properties, which is obviously more powerful than just using LTL formulas in \cite{LS15}. Furthermore, support for continuous time behavior is also possible in our approach in this paper.
\item {\bf Refinement and Equivalence Checking:} In our earlier work \cite{ZHL+17}, equivalence and refinement relations can be proved among different connectors, while the approach in \cite{LS15} is not capable of either equivalence or refinement checking. Furthermore, the constraint solver Z3 is adopted in this work to search for bounded counter examples when the refinement relations cannot be proved in Coq.
\end{itemize}

The paper is organized as follows. After this general introduction, we briefly summarize Reo, Coq and Z3
in Section \ref{sec:pre}. Section \ref{sec:channelandoperator} introduces the
formal model of basic channels, operators and complex connectors in Coq. Section
\ref{sec:verification} shows how to reason about connector properties and equivalence (or
refinement) relations under our Coq formalization. Section \ref{sec:refinement} provides a Reo model in Z3 and shows the usage of Z3 in refinement checking.
In Section \ref{sec:conclusion}, we conclude with some further research directions. Full source codes can be found at \cite{reo2coq2Z3} for further reference.

@@

\subsection{The Coordination Language Reo}
Reo is a channel-based exogenous coordination language wherein complex coordinators, called connectors,
are compositionally built out of simpler ones \cite{Arb04}.
Further details about Reo and its semantics can be found in \cite{Arb04, AR03, BSAR06}.
The simplest connectors are channels with well-defined behavior such as synchronous channels, FIFO channels, etc.
% Each channel in Reo has exactly two directed ends, with their own identities.
There are two types of channel ends: \emph{source} ends and \emph{sink} ends. A source channel end accepts
data into the channel, and a sink channel end dispenses data out of the channel.
\begin{figure}
  \centering
  \begin{tikzpicture}
  \draw (0.7, 0.6) node[draw=none] {Sync};
  \sync{(0,1)}{(1.5,1)}{}
  \draw (2.7, 0.6) node[draw=none] {SyncDrain};
  \syncdrain{(2,1)}{(3.5,1)}{node[below]{}}
  \draw (4.7, 0.6) node[draw=none] {FIFO1};
  \fifoe{(4,1)}{(5.5,1)}{node[below]{}}
  \draw (6.7, 0.6) node[draw=none] {LossySync};
  \lossysync{(6,1)}{(7.5,1)}{node[below left]{}}
  \end{tikzpicture}\vspace{-6mm}
  \caption{Four types of basic channels.}\label{fig:basicchannel}
\end{figure}

The graphical notations of some basic channels are presented in Figure~\ref{fig:basicchannel}, and their behavior can be interpreted as follows:
\begin{itemize}
\item{\textbf{Sync:} a synchronous channel with one source end and one sink end. The pair of I/O operations on its two ends can succeed only simultaneously.}
\item{\textbf{SyncDrain:} a synchronous channel which has two source ends.
 The pair of input operations on its two ends can succeed only simultaneously.
 All data items written to this channel are lost.}
\item{\textbf{FIFO$n$:} an asynchronous channel with one source end and one sink end, and a bounded buffer with capacity $n$.
It can accept data items from its source end. The accepted data items are kept in the internal buffer, and dispensed to
the sink end in FIFO order.
Especially, the FIFO$1$ channel is an instace of FIFO$n$ where the buffer capacity is 1.}
%\item{\textbf{AsyncDrain:}  an asynchronous channel} which has two source ends. The channel guarantees that the operations on its two ends never succeed simultaneously. All data items written to this channel are lost.
\item{\textbf{LossySync:} a synchronous channel with one source end
    and one sink end. The source end always accepts all data items. If
    there is no matching output operation on the sink end of the
    channel at the time that a data item is accepted, then the data
    item is lost; otherwise, the channel transfers the data item
    exactly the same as a Sync channel, and the output operation at the sink end succeeds.}
\end{itemize}

@@

Complex connectors are constructed by composing simpler ones via the \emph{join} and \emph{hiding} operations. Channels are joined together in nodes. The set of channel ends coincident on a node is disjointly partitioned into the sets of source and sink channel ends that coincide on the node, respectively. Nodes are categorized into source, sink and mixed nodes, depending on whether all channel ends that coincide on a node are source ends, sink ends or a combination of the two. The hiding operation is used to hide the internal topology of a connector. The hidden nodes can no longer be accessed or observed from outside. There are three types of operations for channel composition: \emph{flow-through}, \emph{merge} and \emph{replicate}. Figure~\ref{fig:channelcomposition} provides the graphical representation of these operations.

\subsection{Coq and Z3}

Coq \cite{huet1997coq} is a widely-used proof assistant tool, where denotational formalizations (e.g.
theorem and hypothesis) and operational formalizations (e.g. functions and algorithms) are naturally
integrated. Moreover, it allows the interactive construction of formal proofs.
The formal language used in Coq, \emph{Gallina}, provides a convenient way to define
both programming statements and mathematical propositions, for example:
\begin{lstlisting}[language=coq]
(* a variable definition *)
Variables a b: nat.
(* a simple non-recursive function *)
Definition inc(a:nat) := a + 1.
(* axioms don't have to be proved *)
Axiom inc_ax: forall c:nat, inc(c) > c.
(* theorems rely on proving *)
Theorem inc_eq: forall c:nat, inc(c) = c + 1.
Proof.
(* interactive proving based on tactics *)
auto.
Qed.
\end{lstlisting}

@@

Furthermore, Coq is equipped with a set of well-written standard libraries. For example, as used in
this paper, \emph{Stream} provides a co-inductive
definition of infinite lists, and \emph{Reals} defines various operations and theorems on real
numbers. Usually, quite a few lemmas and theorems are pre-defined in such libraries, making it
substantially easier to prove our goals.

@@

Z3 \cite{MouraB08} is an efficient SMT (Satisfiability Modulo Theories) solver freely available from Microsoft. It can be used in various software verification and analysis applications. Acting as a SMT solver, Z3 expands to deciding the satisfiability (or dually the validity) of first order formulas with respect to combinations of theories such as: arithmetic, bit-vectors, arrays, and uninterpreted functions. Given the data and time constraints of connectors, it allows us to decide the satisfiability against properties or refinement relations. Z3 provides bindings for several programming languages. In this paper, we use \emph{Z3 python-bindings} to construct the models and carry out experiments. The following example provides an intuitive understanding of the Z3 solver.
\begin{lstlisting}
x, y = Int('x'), Int('y')

s = Solver()
s.add(x > 10, Or(x + y > 3, x - y < 2))
# Solving constraints in the solver s
print s.check()

# Create a new scope
s.push()
s.add(y < 11)
# Solving updated set of constraints
print s.check()

# Restoring state
s.pop()
# Solving restored set of constraints
print s.check()
\end{lstlisting}

This example contains two constraints at first. From these two constraints, we can see that Z3 supports Boolean operators, such as \emph{And}, \emph{Or}, \emph{Not} and \emph{Implies}, apart from the operators like $<$, $>$ for comparison.
The function \emph{Solver} creates a general purpose solver instance where constraints can be asserted through its \emph{add} method. The method \emph{check} solves the asserted constraints. Finally, the result is satisfiable (unsatisfiable) if a solution is (not) found, respectively. A solver may fail to solve a system of constraints and \emph{unknown} is returned. Each solver maintains a stack of assertions. The command \emph{push} creates a new scope by saving the current stack size. The command \emph{pop} removes all constraints that have been asserted after the last \emph{push} command.

\section{Formal Modeling in Coq}\label{sec:channelandoperator}
In this section, we show how primitive connectors, i.e., channels, and operators for connector composition are specified in Coq and used for modeling of complex connectors. Basic channels, which can be regarded as axioms of the whole framework, are specified as logical predicates illustrating the relation between the timed data streams of input and output. When we need to construct a more complex connector, appropriate composition operators are applied depending on the topological structure of the connector.

\subsection{Basic Definitions}\label{sec:basicdef}

% In this section, we briefly introduce the notion of timed data streams and some pre-defined auxiliary functions and predicates in Coq, which are used in the following sections for modeling connectors.

@@

%\subsection{Auxiliary functions and predicates}
Some auxiliary functions and predicates are defined to facilitate the representation of axioms for basic channels in Reo. This part can be extended for further use in different problems.
The terms \emph{PrL} and \emph{PrR} take a pair of values (a, b) that has Cartesian product type A $\times$ B as the argument and return the first or second value of the pair, respectively.
%\begin{verbatim}
%    Definition PrL {A B: Type}(pair: A * B) := (...).
%    Definition PrR {A B: Type}(pair: A * B) := (...).
%\end{verbatim}
The following functions provide some judgment of time, which can make
the description of axioms and theorems for connectors more concise and
clear: \emph{Teq} means that time of two streams are equal and \emph{Tneq} has the
opposite meaning. \emph{Tle}  (\emph{Tgt}) represents that time of the first stream is
strictly less (greater) than the second stream. The judgement about
equality of data is analogous to the judgement of time. The complete
definition of these functions can be found at \cite{reo2coq2Z3}.

@@

The channel types FIFO and FIFO$n$ where $n$ is an integer greater than $0$ represent the
typical unbounded and bounded asynchronous FIFO channels. A write to a FIFO channel always succeeds, and a write to a FIFO$n$ channel succeeds only if the
number of data items in its buffer is less than its bounded capacity $n$. A read or take operation on a FIFO or FIFO$n$ channel suspends until the first data item
in the channel buffer can be obtained and then the operation succeeds.
%We use the $\overrightarrow{\emph{a}}$ to denote the tail of  a sequence \emph{a}.
For simplicity, we take the FIFO1 channel as an example. This channel type requires that the time when it consumes a data item through its source end is earlier than the time when the data item is delivered through its sink end. Besides, as the buffer has the capacity 1, time of the next data item that flows in should be later than the time when the data in the buffer is delivered. We use intersection of predicates in its definition as follows:
\begin{lstlisting}[language=coq]
Definition FIFO1(Input Output:Stream TD) : Prop :=
    Tle Input Output /\ Tle Output (tl Input) /\ Deq Input Output.
\end{lstlisting}

For a FIFO1 channel whose buffer already contains a data element \emph{e}, the communication can be initiated only if the data element \emph{e} can be taken via the sink end. In this case, the data stream that flows out of the channel should get an extra element \emph{e} settled at the beginning of the stream. And time of the stream that flows into the channel should be earlier than time of the tail of the stream that flows out. But as the buffer contains the data element \emph{e}, new data can be written into the channel only after the element \emph{e} has been taken. Therefore, time of the stream that flows out is earlier than time of the stream that flows in. The channel can be represented as:
\begin{lstlisting}[language=coq]
Definition FIFO1e(Input Output:Stream TD)(e:Data) : Prop :=
    Tgt Input Output /\ Tle Input (tl Output)
    /\ PrR (hd Output) = e  /\ Deq Input (tl Output).
\end{lstlisting}

In the following we choose Axiom to define LossySync
because it is easier to use the coinductive expression to specify
its behavior.

A LossySync channel behaves the same as a Sync channel, except that a write operation on its source end always succeeds immediately. If a compatible read or take operation is already pending on its sink end, the written data item is transferred to the pending operation and both succeed. Otherwise, the write operation succeeds and the data item is lost. The LossySync channel can be defined as follows:
\begin{lstlisting}[language=coq]
Parameter LossySync: Stream TD -> Stream TD -> Prop.
Axiom LossySync_coind:
    forall Input Output: Stream TD,
    LossySync Input Output ->
    (( hd Output = hd Input  /\ LossySync (tl Input)(tl Output)) \/
       LossySync(tl Input) Output).
\end{lstlisting}

% AsyncDrain is analogous to SyncDrain except that it guarantees that the pairs of write operations on the two channel ends never succeed simultaneously. Similarly it only has requirements on the time of the two streams on its opposite ends, but it requires that the times of the two streams are always different. The AsyncDrain channel can be defined as follows:
% \begin{lstlisting}[language=coq]
% Parameter AsyncDrain: Stream TD -> Stream TD -> Prop.
% Axiom AsyncDrain_coind:
%     forall Input1 Input2: Stream TD,
%     AsyncDrain Input1 Input2 ->
%     (~ PrL(hd Input1)  =  PrL (hd Input2) ) /\
%     (
%         ( (PrL(hd Input1) < PrL (hd Input2)) /\ AsyncDrain (tl Input1) Input2) /\
%         ( (PrL(hd Input1) > PrL (hd Input2))  /\ AsyncDrain Input1 (tl Input2) )
%     ).
% \end{lstlisting}

@@

Based on the definition of basic channels and operators, more complex connectors can be constructed structurally.
%As the whole structure of each connector has been determined, the rest work is to use the operators to connect appropriate  channels according to the structure of each connector.
To show how a composite connector is constructed, we consider a simple example as shown in Figure \ref{fig:compsyncfifo}, where a FIFO1 channel is attached
to the sink end of a Sync channel. Assume \emph{AB} is of type Sync and \emph{BC} is of type FIFO1, then we can construct the required connector by
illustrating \emph{Sync(A,B)} and \emph{FIFO1(B,C)}. The configuration and the functionality of the required connector can be specified using this concise method. Note that the composition operations can be easily generalized to the case of multiple nodes, where the modeling of connectors is similar. More
examples can be found in Section \ref{sec:verification}.

\section{Reasoning about Connectors in Coq}\label{sec:verification}
After modeling a connector in Coq, we can analyse and prove important properties of the connector. In this section, we give some examples to elucidate how to
reason about connector properties and prove refinement/equivalence relations between different connectors, with the help of Coq.

\subsection{Derivation of Connector Properties}
The proof process of a property is as follows: the user states the proposition that needs to be proved, called a \emph{goal},
then he/she applies commands called \emph{tactics} to decompose this goal into simpler subgoals or solve it directly. This decomposition
process ends when all subgoals are completely solved. In the following, we use some examples to illustrate our approach instead of
giving all the complex technical details.
\begin{example}
We first consider the connector given in Figure \ref{fig:compsyncfifo}, which consists of two channels \emph{AB} and \emph{BC} with types Sync and FIFO1, respectively.
We use $a$ and $b$ to denote the time streams when the corresponding data streams flow into and out of the Sync channel \emph{AB}, and
$c$ to denote the time stream for the data stream that flows out of the FIFO1 channel \emph{BC}. Here we can see that a flow-through
operation has acted on the mixed node \emph{B}. The time when the stream flows into the FIFO1 channel \emph{BC} is equal to the time when the
stream flows out of the Sync channel \emph{AB}. The following theorem states the property $a < c$ for this connector. The connector is based on the
axioms Sync and FIFO1, which can be used as hypotheses for the proof of the following theorem.
\end{example}

\begin{theorem}[Transfer\_le]\label{the:tleac}
$\forall A,B,C.\:Sync(A,B)\land FIFO1(B,C) \rightarrow Tle(A,C)$.
\end{theorem}

In Coq, the theorem can be proved as follows:
\begin{lstlisting}[language=coq]
Theorem thm_4_1: forall A B C,
  Sync A B /\ FIFO1 B C -> Tle A C.
Proof.
    intros. destruct H. destruct H0.
    intro n. rewrite H. apply H0.
Qed.
\end{lstlisting}
